---
title: "Data607-MajorAssignment-Project2-Data Transformation"
author: "Vinayak Kamath"
date: "3/7/2020"
output:
  html_document:
    df_print: paged
    toc: false
    toc_depth: 3
    toc_float: 
      collapsed: true
      smooth_scroll: true
      number_sections: true
    theme: cosmo
  pdf_document:
    extra_dependencies:
    - geometry
    - multicol
---

# Data Transformation {.tabset}  

Below three of the “wide” datasets identified in the Week 6 Discussion items  have been used for this exercise.  
  
1. Set 1  - Bank stocks from 2007 (*Discussion Thread by Jeff Shamp*)  
2. Set 2  - UNICEF dataset on Under 5 Mortality (*Discussion Thread by Samuel Bellows*)   
3. Set 3  - Hospital Consumer Assessment of Healthcare Providers and Systems response data by state (*Discussion Thread by Thomas Hill*)  
  
  
We will practice tidying and transformations on these data sets.and We will performing few analysis points as discussed/requested in the discussion threads. 


```{r setup, include=FALSE, echo=FALSE }
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(stringr)
library(maptools)
```



## Bank Stocks  
Reading the CSV file from GIT repository and loading into dataframe:  
```{r, warning = FALSE}
theUrl <- "https://raw.githubusercontent.com/kamathvk1982/Data607-MajorAssignment-Project2/master/banks.csv"
banks.full.df <- read.csv(file = theUrl, header = T , sep = ',', na.strings=c("NA","NaN", "") )

# Creating new data frame with reduced columns for current analysis:
banks.df <-  banks.full.df %>%
  select(c(date=Bank.Ticker, date.for.split=Bank.Ticker, bac.close=BAC.3, bac.volume=BAC.4, c.close=C.3, c.volume=C.4, jpm.close=JPM.3, jpm.volume=JPM.4, gs.close=GS.3, gs.volume=GS.4)) %>%
  filter(date!= 'Stock Info' ,  date!= 'Date'  ) %>%
  separate(date.for.split,  c("date.year", "date.month")) %>%
  unite("date.year.month", date.year:date.month, sep='-')

kable(head(banks.df))

# Next, we will tidy the data by reshaping the data layput in the table by using tidyr->gather function:  
banks.tidy.df <- gather(banks.df, key = "key", value = "value", bac.close , bac.volume , c.close , c.volume , jpm.close , jpm.volume , gs.close , gs.volume )

banks.tidy.df$value  <- as.numeric(as.character(banks.tidy.df$value))

kable(head(banks.tidy.df))
```  
  
***  

#### **Analysis 1**  
Piloting the closing balance and volume traded of each bank using point chart:  
```{r, warning = FALSE}
plot.price <- banks.tidy.df %>%
  filter(grepl('close', key) ) %>%
   group_by(key,date.year.month) %>%
     summarize(mean.price= mean(as.double(value)))  %>%
       ggplot(aes(x=date.year.month, y=mean.price, colour=key)) + 
  theme(axis.text.x = element_text(angle = 90, size = 2)) +
  geom_point() 

plot.volume <- banks.tidy.df %>%
  filter(grepl('volume', key) ) %>%
   group_by(key,date.year.month) %>%
     summarize(mean.volume= mean(as.double(value)))  %>%
       ggplot(aes(x=date.year.month, y=log(mean.volume), colour=key)) + 
  theme(axis.text.x = element_text(angle = 90, size = 2)) +
  geom_point()    

plot.price
plot.volume
```  
  
***Based on the above point chart we can see how the stock prices for banks have been impacted during recession.***  
  
***  
  
#### **Analysis 2**  
Comparing for Citi and JP Morgan; Getting the mean of the prices for the year 2008; the peak of the recession:  
```{r, warning = FALSE}
data.citi.2008 <- banks.tidy.df %>%
  filter(grepl('c.close', key) , grepl('2008', date.year.month)) %>%
   separate(date.year.month,  c("date.year", "date.month"))  

data.jpm.2008 <- banks.tidy.df %>%
  filter(grepl('jpm.close', key) , grepl('2008', date.year.month)) %>%
   separate(date.year.month,  c("date.year", "date.month"))  

#52 week data for Citi for 2008
summary(as.double(data.citi.2008$value))
boxplot(as.double(data.citi.2008$value), main="CITI 2008", col = "green")

#52 week data for JPM for 2008
summary(as.double(data.jpm.2008$value))
boxplot(as.double(data.jpm.2008$value), main="JPM 2008", col = "purple")
```  

***Based on the above box plot shapes we can say that Citi had a much bigger or higher price dip then JPM.***  

   
***  

## UNICEF  

Reading the CSV file from GIT repository and loading into dataframe:  
```{r, warning = FALSE}
theUrl <- "https://raw.githubusercontent.com/kamathvk1982/Data607-MajorAssignment-Project2/master/unicef-u5mr.csv"
unicef.full.df <- read.csv(file = theUrl, header = T , sep = ',', na.strings=c("NA","NaN", "") )
dim((unicef.full.df))
```

Data Transformation and Tidy using dplyr and tidyr: 
```{r}
# Next, we will tidy the data by reshaping the data layput in the table by using tidyr->gather function:
unicef.tidy.df <- gather(unicef.full.df, key = "Year", value = "Value", -CountryName) 
unicef.tidy.df$Value  <- as.numeric(as.character(unicef.tidy.df$Value))
unicef.tidy.df$CountryName  <- str_trim(as.character(unicef.tidy.df$CountryName))

# Use the tidyr->drop_na function to drop the row on column Status having NA value:
unicef.tidy.df <- drop_na(unicef.tidy.df, Value)    

# Use the sub  function to drop the 'U5MR.' from new column Year:
unicef.tidy.df$Year  <- sub('U5MR.','',unicef.tidy.df$Year)

kable(head(unicef.tidy.df))

```  
  
#### **Analysis 1**  
Comparing a Developed Nation **United States of America** , a Developing nation **India** and a Under Developed Nation **Chad**:  
```{r}
unicef.set1.df <-  unicef.tidy.df %>%
  filter(grepl('United States of America|India|Chad' , CountryName	)) %>%
  arrange(Year, CountryName)

kable(head(unicef.set1.df))

unicef.set1.df %>%
       ggplot(aes(x=Year, y=Value, colour=CountryName)) + 
  theme(axis.text.x = element_text(angle = 90, size = 2)) +
  geom_point() 

```  

***Based on the above Chart, we can see that the under 5 mortality is coming down for all three countries; but comparatively the counts are still high for Under Developed Countries and for Developing Countries .***  

***  

#### **Analysis 2**  
List of countries where the under 5 mortality  as of 2015 is still greater then 85:    
```{r}
# Filtering for 2015 and greater then 85:
unicef.set2.df <-  unicef.tidy.df %>%
  filter( Year=='2015' ,  Value > 85 ) %>%
  arrange( desc(Value))


# Ploting on world map:
data(wrld_simpl)
myCountries = wrld_simpl@data$NAME %in% names(table(unicef.set2.df$CountryName))
plot(wrld_simpl, col = c(gray(.80), "red")[myCountries+1])
```  

***Based on the above World Map plotting we can see that most of these countries are in Continent Africa.***  

***  
  
  
## Hospital Consumer Assessment  

Reading the CSV file from GIT repository and loading into dataframe:  
```{r, warning = FALSE}
theUrl <- "https://raw.githubusercontent.com/kamathvk1982/Data607-MajorAssignment-Project2/master/HCAHPS.csv"
hcahps.full.df <- read.csv(file = theUrl, header = T , sep = ',', na.strings=c("NA","NaN", "Not Available") )

hcahps.full.df$HCAHPS.Answer.Percent  <- as.numeric(as.character(hcahps.full.df$HCAHPS.Answer.Percent))

kable(head(hcahps.full.df))
```
  
***  
  
#### **Analysis 1**  
Measure RESPECT (treated patients with courtesy and respect) for Nurses and Doctors for NJ and near by States:  
```{r, warning = FALSE}
# we will create the required dataset using select, filter and separate function for data transformation:
respect.df <-  hcahps.full.df %>%
  select(c(State, Measure.ID = HCAHPS.Measure.ID,Answer.Percent=HCAHPS.Answer.Percent )) %>%
  filter(grepl('CT|NY|PA|NJ' , State	) ,  grepl('RESPECT' , Measure.ID	) ) %>%
  separate(Measure.ID,  c("Type", "Response"), sep = '_RESPECT_')


# Next, we will tidy the data by reshaping the data layput in the table by using tidyr->spread function:
respect.tidy.df <- spread(respect.df, key = Response, value = Answer.Percent  )
colnames(respect.tidy.df) <- c("State", "Type", "Always", "Sometimes.or.Never", "Usually")

kable(respect.tidy.df)

respect.tidy.df %>% group_by(State, Type) %>% summarise (Positive.Ind = sum(as.integer(Always)+as.integer(Usually)))

```  

***If we treat "Usually" and "Always" as a positive indicator then we can say that in State of CT the Doctors did better and in the State of PA the Nurses did better; for NJ and NY state the Doctors and Nurses were voted equally.***   
  
***  

#### **Analysis 2**   
Measure RATING (a rating of 9 or 10 on a scale from 0 (lowest) to 10 (highest)) for HOSPITALS for NJ and near by States:  
```{r, warning = FALSE}
# we will create the required dataset using select, filter and separate function for data transformation:
rating.df <-  hcahps.full.df %>%
  select(c(State, Measure.ID = HCAHPS.Measure.ID,Answer.Percent=HCAHPS.Answer.Percent )) %>%
  filter(grepl('CT|NY|PA|NJ' , State	) ,  grepl('RATING' , Measure.ID	) ) %>%
  separate(Measure.ID,  c("Type", "Response"), sep = '_RATING_')

# Next, we will tidy the data by reshaping the data layput in the table by using tidyr->spread function:
rating.tidy.df <- spread(rating.df, key = Response, value = Answer.Percent  )
colnames(rating.tidy.df) <- c("State", "Type", "low.6.or.Lower", "medium.7.or.8", "high.9.or.10")

kable(rating.tidy.df)

rating.tidy.df %>% group_by(State, Type) %>% summarise (Positive.Ind = sum(as.integer(medium.7.or.8)+as.integer(high.9.or.10)))

```  

***If we treat "rating of 7 or 8 [medium]" and "rating of 9 or 10 [high]" as a positive indicator then we can say that in State of PA the Hospitals did better then other states.***   
  
***  

#### **Analysis 3**
Measure Doctors and Nurses across applicable questions in all States:  
```{r, warning = FALSE}
# we will create the required dataset using select, filter and separate function for data transformation.
# BY looking at the data values we can say that in column Measure.ID any pattern of DOCTOR or COMP_2 can be treated as DOCTOR data
# and any pattern of NURSE or COMP_1 can be treated as NURSE data:  

compare.doctor.df <-  hcahps.full.df %>%
  select(c(State, Measure.ID = HCAHPS.Measure.ID,Measure.ID2 = HCAHPS.Measure.ID, Answer.Percent=HCAHPS.Answer.Percent )) %>%
  filter(   grepl('DOCTOR|COMP_2' , Measure.ID	), grepl('_A_P|_U_P' , Measure.ID2	) ) 

compare.nurse.df <-  hcahps.full.df %>%
  select(c(State, Measure.ID = HCAHPS.Measure.ID,Measure.ID2 = HCAHPS.Measure.ID, Answer.Percent=HCAHPS.Answer.Percent )) %>%
  filter(   grepl('NURSE|COMP_1' , Measure.ID	), grepl('_A_P|_U_P' , Measure.ID2	) ) 
```

Next, we will calculate the sum of the scores:  
```{r echo=FALSE}
paste('Total of Positive Response Score for Doctors is', sum(as.double(compare.doctor.df$Answer.Percent), na.rm = T) )
paste('Total of Positive Response Score for Nurses is', sum(as.double(compare.nurse.df$Answer.Percent), na.rm = T) )
```  

***If we treat "Usually [_U_P]" and "Always [_A_P]" as a positive indicator then we can say that the Nurses had a better result then Doctors across all states***   
  
***    
